### 2020-02-20 Common/chapters/deep_learning.tex
Julien and Johan: typo
Johan and Julien: Corrections according to Akhansha's feedback. All done except:  - Logo p.1 - currently one of the trendiest topics p.9 - activating invasion and metastasis p.35

\chapter{Deep learning}
\label{ch:deep_learning}
This chapter provides the theoretical foundation in deep learning which is required to understand of the rest of the work. It starts with an historical timeline of deep learning before describing what a neural network is. Then, the notion of training a neural network, with all that is involved such as forward propagation, backpropagation, hyperparameters, data splitting or performance evaluation is explained. This part is followed by another section devoted to a special type of neural networks, the "convolutional neural networks". They are used in many computer vision applications due to their great performance for these tasks. Finally, the concept of "transfer learning" is discussed since the entire chapter \ref{ch:transfer_learning} relies on it.

\section{Introduction to deep learning}
Deep learning is currently one of the trendiest topics in machine learning, a subset of artificial intelligence. Machine learning refers to statistical models that allow computers to perform specific tasks without having been explicitly programmed to solve them. In fact, these models try to find structural patterns within data in order to understand new incoming situations and react in the best possible way. There exist various techniques in machine learning such as k-NN, SVM, k-means, decision trees, association rules, etc. What mainly differentiates deep learning from these algorithms is the concept of neural networks (see section  \ref{what_is_a_neural_network}) that are combined to form deep neural networks.\\
Neural networks are inspired from the biological neural networks of the brain. These systems try to learn how to solve a problem based on the data they receive as input. Many concrete applications make use of neural networks: autonomous vehicles, smarter translators, computer-aided diagnoses, personal assistants, art creation, robotics, etc. The presence of deep learning techniques in these use cases clearly testifies to the enthusiasm of many areas for this technology. Furthermore, since this field has recently gained interest (see section \ref{historical_background}), a lot of research is still ongoing, which suggests that many exciting new applications will certainly be discovered in the near future.

\subsection{Historical background}
\label{historical_background}
As described on figure \ref{history}, the theoretical foundations of deep learning appeared long before the invention of computers. From the first attempts to understand the human brain until today, huge progress was made to be establish the basic components of modern neural networks. One could ask why deep learning took off recently if the theory was around for a long time.\\
As stated by Goodfellow et al. \cite{15}, the first part of the answer is computing power. In fact, deep learning algorithms need a lot of data to work properly, which requires powerful CPUs/GPUs that either didn't exist or were only within few people's reach. One other main reason concerns the lack of data. Since deep learning algorithms "learn" from data, learning is impossible if good-quality data are not available. The era of Big Data enhanced deep learning possibilities. 
%%%% Je mettrais soit l'un soit l'autre. Le paragraphe du dessus c'est le paragraphe du dessous avec des autres mots :)
% These two points are summarised: "The increase in model size overtime, due to the availability of faster CPUs, the advent of general purpose GPUs, fasternetwork, connectivity and better software infrastructure for distributed computing, is one of the most important trends in the history of deep learning. This trend is generally expected to continue well into the future".\\
Finally, before the year 2012, the abilities of neural networks were still to be proven. This changed with the ImageNet Large Scale Visual Recognition Challenge (a competition where researchers evaluate their algorithms on several visual recognition tasks). In fact, the deep convolutional neural network called "AlexNet" achieved 16\% of classification error rate, whereas the previous best scores were around 25\%. This victory marked the beginning of a new craze for these types of algorithms.

@@

From a descriptive point of view, neural networks can simply be seen as a non-linear applications that associate an input to an output with respect to certain parameters. The input can be an image, a sound or any input that can be converted into numerical features. The output of a neural network depends on the problem it tries to solve. In computer vision, the most common types of outputs are classes (for classification problems) and pixel coordinates (for segmentation problems).\\
From a mathematical standpoint, a neural network can be defined as a non-linear function $f$ that associates to an input $x$ an output $y$ with respect to parameters $\theta$.
\begin{equation}
y = f(x,\theta)
\end{equation}
The parameters $\theta$ are estimated from the training samples.
%\begin{figure}[!h]
%\centering
%\includegraphics[width=1\textwidth, keepaspectratio=true]{./figures/neural_network.png}
%\caption{Neural Network example}
%\label{neural_network}
%\end{figure}

\subsection{Supervised and unsupervised learning}
In machine learning, two different kinds of tasks exist. The first one is "supervised learning". It includes learning algorithms whose training samples are associated to their labels in order to find the optimal mapping between the input and the output. The second one is "unsupervised learning". In contrast to supervised algorithms, the latter rely on unlabeled data. Its main goal is to infer the natural structure  present in the data. Since the models presented in this work belong to the "supervised learning" category, notions explained below refer to this kind of algorithms.

@@

\noindent \textbf{General comment}
\begin{itemize}
\item Superscript (i) denotes the $i^{th}$ training example.
\item Superscript [l] denotes the $l^{th}$ layer of the neural network.
\end{itemize}

@@

\noindent \textbf{Neural networks components}
\begin{itemize}
\item $X\in \R$ is the input matrix matrix of a neural network.
\item $x^{(i)} \in \R^{n_{x}}$ is the $i^{th}$ example (sample) represented as a column vector.
\item $Y \in \R^{n_{y} \times m}$ is the label matrix.
\item $y^{(i) \in \R^{n_{y}}}$ is the output label for the $i^{th}$ example.
\item $W^{[l]} \in \R$ \textsuperscript{\# of neurons in the next layer = j  x \# of neurons in the previous layer = k} is the weight matrix at layer $[l]$.
\item $b^{[l]} \in R$\textsuperscript{\# of units in next layer} is the bias vector in the $l^{th}$ layer.
\item $\hat{y} \in R^{n_{y}}$ is the predicted output vector. It can also be denoted $a^{[L]}$ where $L$ is the number of layers in the whole network.
\item $g^{[l]}(x)$ is the $l^{th}$ activation function.  
\item $z^{[l]} = W_{x}x^{(i)} + b^{[l]}$ denotes the weighted sum of the input given to the $l^{th}$ layer before passing through the activation function.\\
\end{itemize}

@@

\subsection{Perceptrons}
\label{perceptron}
Perceptrons are the main components of neural networks. They were "developed in the 1950s and 1960s by the scientist Frank Rosenblatt, inspired by earlier work by Warren McCulloch and Walter Pitts" \cite{13}. Today, they are called "artificial neurons" or simply "neurons".\\
A perceptron $j$ is a function $f$ of input $x=(x_{1}, ..., x_{n})$ weighted by a vector of weights $w_{}=(w_{1}, ..., w_{n})$, completed by a bias $b_{j}$ and associated to a non-linear activation function $g$:
\begin{equation}
\label{perceptron_equation}
a_{j} = f_{j}(x) = g((\sum_{k=1}^{n} x_{k} * w_{k}) + b_{})
\end{equation}

Schematically speaking, a perceptron can be represented as on figure \ref{perceptron_model}. Each input is multiplied with its corresponding weight. The sum of this result then goes through a non-linear function, called "activation function". This activation function acts like a threshold that determines the proportion of the result that goes further in the network. There exist multiple activation functions (see \ref{activation_functions}). It is extremely important to use non-linear functions instead of a linear functions. In fact, the output of a perceptron is given as input to the others (see \ref{multilayer_perceptron}). Consequently, if linear functions only are used throughout the network, linear outputs are given as inputs to other linear functions. Since the composition of two linear functions is itself a linear function, assembling perceptrons to create neural networks of multiple layers would not make sense anymore.

@@

\subsection{Multilayer perceptrons}
\label{multilayer_perceptron}
A multilayer perceptron is a type of artifical neural networks. Du et al. \cite{23} define multilayer perceptrons as "feedforward networks with one or more layers of units between the
input and output layers" where "the output units represent a hyperplane in the space
of the input patterns."\\
A multilayer perceptron is composed of $L$ layers, each of them composed of $n^{l}_{h}$ perceptrons. The layers are organized in the following way:
\begin{itemize}
\item The input layer: it is the neural network entry point for the data. Generally speaking, the data is provided in the form of a matrix $X \in \R$ of size $(n_{x} \times batch\_size)$ with their corresponding labels $Y \in \R$ of size $(n_{y} \times batch\_size)$. The batch size defines the number of samples that will be given at the same time to the network and $n_{x}$ is the dimension of each sample. Moreover, $x^{(i)}$ is the $i^{th}$ sample represented as a column vector. The total number of training samples is given by $m$. Finally, $y^{(i)}$ is the output label for the $i^{th}$ example. For instance, suppose the number of samples is 100 and the batch size 32. In this situation, the network will be feeded with 4 batches of sizes [32, 32, 32, 4] respectively.

\item The hidden layer(s): hidden layers stand for all layers that are between the input layer and the output layer. Each of them has its own weights and biases (W, b), denoted by $W^{[l]} \in \R $ and $b^{[l]}$ respectively, where $W_{ij}^{l}$ corresponds to the weights associated with the connection between perceptron $j$ in layer $l$ and perceptron $i$ in layer $l+1$. By analogy, $b_{i}^{[l]}$ is the bias associated with perceptron $i$ in layer $l$. Weights and biases are the parameters to optimize in order to obtain the best mapping between the input and the output of the network (see section \ref{training_a_neural_network} ). Before training the neural network, the weights can be randomly initialized or initialized with more sophisticated methods such as "Xavier initialization" or "Kaiming initialization"(see section \ref{weight_initialization}).

\item The output layer: it is the last layer of the neural network. Its role is essential since it produces the prediction of the network for a given input. The prediction of a neural network is given by $\hat{y} \in R^{n_{y}}$ with $n_{y}$ beeing the number of different labels. In a classification task, whose goal is to assign to each input a specific class, the $\hat{y}$ is usually the probability that the input belongs to each class. In that case, the \textit{softmax} activation function would be used at the end.
\end{itemize}
The advantage of organizing the weights, biases and inputs in matrices is due to the ability of modern CPUs/GPUs to quickly perform linear algebra computations. This way of structuring the network component is called "vectorization" which avoids using loops in the code, which would considerably slow down the computations.
Figure \ref{multilayer_perceptrons_figure} illustrates the concept of multilayer perceptrons. In this example, the total number of layers $L$ is equal to 3, the input size $n_{x}$ is equal to 4 and the number of units in in each layer is $n_{h}^{1}= 4$, $n_{h}^{2}=2$, $n_{h}^{3}=1$. The network contains the weights  $W^{(1)} \in \R^{2x3}$, $W^{(2)} \in \R^{1x2}$ and the biases  $b^{[1]}=2$, $b^{[2]}=1$.

@@

In other words, after each forward pass through the entire network, backpropagation performs a backward pass which aims at minimizing the cost function by adjusting parameters of the model. The way parameters are updated is defined by the gradients of the cost function with respect to each parameter of the network. The gradient of the cost function $J(x_{1}, x_{2}, ..., x_{m})$ at point $x$ is given by:
\begin{equation}
\frac{\partial J}{\partial x} = [\frac{\partial J}{\partial x_{1}}, \frac{\partial J}{\partial x_{2}}, ..., \frac{\partial J}{\partial x_{1}}]
\end{equation}
The gradient shows how much all parameters that constitute x need to change to minimize the function. In neural networks, the parameters of the cost function are all weight matrices $W^{[l]}$ and biases $b^{[l]}$. The computation of all these gradients relies on the "chain rule". Inthe case of weights, the chain rule is:
\begin{equation}
\frac{\partial J}{\partial w_{jk}^{l}} = \frac{\partial J}{\partial z_{j}^{l}} * \frac{\partial z_{j}^{l}}{\partial w_{jk}^{l}}
\end{equation}
Similarly, the chain rules has to be applied to the biases:
\begin{equation}
\frac{\partial J}{\partial b_{j}^{l}} = \frac{\partial J}{\partial z_{j}^{l}} * \frac{\partial z_{j}^{l}}{\partial b_{j}^{l}}
\end{equation}
Once the gradients of each parameter are computed, these parameters are updated. The weights update is described by the following equation:
\begin{equation}
W^{[l]} = W^{[l]} - \alpha * \frac{\partial J}{\partial W^{[l]}}
\end{equation}
The biases update corresponds to:
\begin{equation}
b^{[l]} = b^{[l]} - \alpha * \frac{\partial J}{\partial b^{[l]}}
\end{equation}
%The hyperparameter $\alpha$ of these two equations is called "learning rate". It determines the gradient's influence at each epoch and has to be manually tuned.
The "learning rate" $\alpha$ determines the influence that the gradient has at each epoch. It is an hyperparameter and has to be tuned manually.

@@

\subsection{Data}
In deep learning, data is essential. As seen previously, neural networks learn features from it. Therefore, data has to be handled carefully and in the right way. Usually, it is split into three different sets:
\begin{itemize}
\item The training set: this is the part of the dataset that is used to train the neural network (the weights and biases).
\item The validation set: this the dataset that is used to evaluate a trained model. Usually, the evaluation on the validation set is performed every $N$ epochs, where $N$ is a fixed number. The validation set needs to come from the same distribution as the training set but should contain exclusively unseen data. This last point is crucial since the validation set shows how well the neural network generalizes on unknown data. The validation set can also be used as indicator to decide when the training should be stopped in order to prevent "overfitting" which is the behaviour of a model that fits to the training set too closely and don't generalize well. In fact, if the validation loss continuously increases for a certain number of epochs, going on with the training will increase the overfit. This fact of interrupting the training earlier is called "early stopping".
\item The test set: this last part of the dataset is used to establish the final evaluation of the model. It also contains unseen data exclusively. 
\end{itemize}
Regarding the way these three sets are split, it mostly depends on the number of samples available. If the latter is big, the data is split into training and testing using the ratio 80/20. Then the remaining training samples are also split into training and validation using the ratio 80/20. On the contrary, if there are few data available, k-fold cross-validation is a good practice. This technique consists in splitting the entire dataset into k folds. One fold is picked as test set and the others are considered as training sets. The model is trained on training folds and tested on the test set. Then, another test set is picked and the same process is repeated until all possible tests set are picked. At the end of the process, the results of all test sets are averaged, which provides a good estimation of the model's performance. This technique is summarized on figure \ref{cross_validation}.

@@

\subsection{Hyperparameters tuning}
Hyperparameters denote parameters that cannot be directly learned from the data. That is the case for the learning rate, the batch size and the number of epochs that were described in previous sections. So, these parameters have to be manually tuned in order to find the best configuration (i.e the one that minimizes the cost function and that keeps a acceptable level of generalization).\\
Regarding the learning rate $\alpha$, its value has to be neither too large or too small. A too large learning rate is recognizable by analyzing the training loss curve: if "the loss is exploding or fluctuating" or if "it has stopped improving and it is wandering around a suboptimal local optima" \cite{28}, the learning rate is too high and should be decreased. On the contrary, if "the learning is very slow and the loss is decreasing consistently" or if "the model is overfitting" \cite{28}, it is a clear sign that it should be increased.
The learning rate can take a wide range of values. Consequently, the most used technique to find the optimal learning rate is simply the "trial and error", which consists in "trying widely different learning rates to determine the range of learning rates that need to be explored" \cite{28}. There also exists methods that, instead of keeping a fixed learning rate for the entire training, reduce it after each epoch (learning rate decay) or each time the loss on the validation set does not decrease (learning rate scheduling).\\
Batch size is another important hyperparameter to tune. Training a network with a small batch size "requires less memory, since the latter is trained using fewer samples" \cite{29}. Furthermore, "networks train faster because the weights update is done after each propagation" \cite{29}. Nevertheless, "the smaller the batch the less accurate the estimate of the gradient will be" \cite{29}. Indeed, due to the high weights update frequency, the gradient fluctuates much more than if was computed after a bigger number of samples.\\
Finally, the number of epochs during which the network is trained has to be carefully chosen. In fact, from a certain point in the training, neural networks don't learn anymore useful features in the data and start overfitting. This point corresponds to the moment where, the validation loss does not decrease anymore and starts to increase continuously. It is exactly at this moment that the training should stop. To achieve this goal, it is possible either to directly choose the right number of epochs or to use "early stopping", which stops the training as soon as the validation loss does not decrease for $N$ epochs.

@@

To perform a convolution, numerical input data and a "filter". A filter can be seen as a an $fxf$ numerical patch that moves across the entire input. It first moves horizontally until reaching the right-most border of the image. Then, it goes down a cell and starts from the border on the left-hand side. This process is repeated until the filter reaches the bottom-right corner of the image, which marks the end of the convolution. At each step, the dot product between the filter and the part of the input covered by the filter is computed. Figure \ref{fig:convolution} showcases a simple convolution which aims at finding vertical lines in a black and white image. For example, the output of the first step of the convolution (in red) is computed by evaluating the dot product between the blue filter and the red part of the input image: 
\begin{equation}
\begin{gathered}
\color{blue}(-1) \color{black}* \color{red}0 \color{black}+ \color{blue}2 \color{black}* \color{red}0 \color{black}+ \color{blue}(-1) \color{black}* \color{red}0
\\ \color{black}+ \\ 
\color{blue}(-1) \color{black}* \color{red}0 \color{black}+ \color{blue}2 \color{black}* \color{red}0 \color{black}+ \color{blue}(-1) \color{black}* \color{red}1 
\\ \color{black}+ \\ 
\color{blue}(-1) \color{black}* \color{red}0 \color{black}+ \color{blue}2 \color{black}* \color{red}0 \color{black}+ \color{blue}(-1) \color{black}* \color{red}1 
\\ \color{black}= -2
\end{gathered}
\end{equation}

@@

Different parameters can change the way a convolution behaves. First of all, the previous example relied on a filter moving by respectively one cell to the right and to the bottom. In this case, the so-called "stride" is equal to 1. Other applications could rely on a bigger stride. Furthermore, the previous example reduced the output size of $3x3$ in proportion to the initial input size of $5x5$. To influence the output size, a padding can be added to the outside of the input image, usually filled with 0s. Three ways of padding images are commonly used as shown on figure \ref{fig:convolution_padding}:
\begin{itemize}
	\item \textbf{Valid:} the input image is not padded. This means that the filter only goes through existing pixel values, which makes the output size smaller than the input size.

	\item \textbf{Same:} the input image is padded in a way that makes the output size the same as the input size.

	\item \textbf{Full:} the input image is padded so that, in the first step of the convolution, only the bottom-right cell of the filter overlays the pixel value of the image, the rest overlaying padding cells. This makes the output size larger than the input size. 
\end{itemize}

>>>>

\chapter{Deep learning}
\label{ch:deep_learning}
This chapter provides the theoretical foundation in deep learning which is required to understand the rest of the work. It starts with a historical timeline of deep learning before describing what a neural network is. Then, the notion of training a neural network, with all that is involved such as forward propagation, backpropagation, hyperparameters, data splitting or performance evaluation is explained. This part is followed by another section devoted to a special type of neural networks, the "convolutional neural networks". They are used in many computer vision applications due to their great performance on these tasks. Finally, the concept of "transfer learning" is discussed as the entire chapter \ref{ch:transfer_learning} relies on it.

\section{Introduction to deep learning}
Deep learning is currently one of the trendiest topics in machine learning, a subset of artificial intelligence. Machine learning refers to statistical models that allow computers to perform specific tasks without having been explicitly programmed to solve them. In fact, these models try to find structural patterns within data in order to understand new incoming situations and react in the best possible way. There exist various techniques in machine learning such as k-NN, SVM, k-means, decision trees, association rules, etc. What mainly differentiates deep learning from these algorithms is the concept of neural networks (see Section  \ref{what_is_a_neural_network}) that are combined to form deep neural networks.\\
Neural networks are inspired from the biological neural networks of the brain. These systems try to learn how to solve a problem based on the data they receive as input. Many concrete applications make use of neural networks: autonomous vehicles, smarter translators, computer-aided diagnoses, personal assistants, art creation, robotics etc. The presence of deep learning techniques in these use cases clearly testifies to the enthusiasm of many areas for this technology. Furthermore, as this field has recently gained interest (see Section \ref{historical_background}), a lot of research is still ongoing, which suggests that many exciting new applications will certainly be discovered in the near future.

\subsection{Historical background}
\label{historical_background}
As described on figure \ref{history}, the theoretical foundations of deep learning appeared long before the invention of computers. From the first attempts to understand the human brain until today, huge progress was made to establish the basic components of modern neural networks. One could ask why deep learning took off recently if the theory was around for a long time.\\
As stated by Goodfellow et al. \cite{15}, the first part of the answer is computing power. In fact, deep learning algorithms need a lot of data to work properly, which requires powerful CPUs/GPUs that either didn't exist or were only within few people's reach. One other main reason concerns the lack of data. Since deep learning algorithms "learn" from data, learning is impossible if good-quality data is not available. The era of Big Data enhanced deep learning possibilities. 
%%%% Je mettrais soit l'un soit l'autre. Le paragraphe du dessus c'est le paragraphe du dessous avec des autres mots :)
% These two points are summarised: "The increase in model size overtime, due to the availability of faster CPUs, the advent of general purpose GPUs, fasternetwork, connectivity and better software infrastructure for distributed computing, is one of the most important trends in the history of deep learning. This trend is generally expected to continue well into the future".\\
Finally, before the year 2012, the abilities of neural networks were still to be proven. This changed with the ImageNet Large Scale Visual Recognition Challenge (a competition where researchers evaluate their algorithms on several visual recognition tasks). In fact, the deep convolutional neural network called "AlexNet" achieved 16\% of classification error rate, whereas the previous best scores were around 25\%. This victory marked the beginning of a new craze for these types of algorithms.

@@

From a descriptive point of view, neural networks can simply be seen as non-linear applications that associate an input to an output with respect to certain parameters. The input can be an image, a sound or any input that can be converted into numerical features. The output of a neural network depends on the problem it tries to solve. In computer vision, the most common types of outputs are classes (for classification problems) and pixel coordinates (for segmentation problems).\\
From a mathematical standpoint, a neural network can be defined as a non-linear function $f$ that associates an input $x$ to an output $y$ with respect to parameters $\theta$.
\begin{equation}
y = f(x,\theta)
\end{equation}
The parameters $\theta$ are estimated from the training samples.
%\begin{figure}[!h]
%\centering
%\includegraphics[width=1\textwidth, keepaspectratio=true]{./figures/neural_network.png}
%\caption{Neural Network example}
%\label{neural_network}
%\end{figure}

\subsection{Supervised and unsupervised learning}
Machine learning algorithms belongs can be classified in two classes. The first one is "supervised learning". It includes learning algorithms whose training samples are associated with their labels in order to find the optimal mapping between the input and the output. The second one is "unsupervised learning". In contrast to supervised algorithms, the latter rely on unlabeled data. Its main goal is to infer the natural structure  present in the data. Since the models presented in this work belong to the "supervised learning" category, notions explained below refer to this kind of algorithms.

@@

\noindent \textbf{General comment}
\begin{itemize}
\item Superscript ($i)$ denotes the $i^{th}$ training example.
\item Superscript $[l]$ denotes the $l^{th}$ layer of the neural network.
\end{itemize}

@@

\noindent \textbf{Neural networks components}
\begin{itemize}
\item $X\in \R$ is the input matrix of a neural network.
\item $x^{(i)} \in \R^{n_{x}}$ is the $i^{th}$ example (sample) represented as a column vector.
\item $Y \in \R^{n_{y} \times m}$ is the label matrix.
\item $y^{(i) \in \R^{n_{y}}}$ is the output label for the $i^{th}$ example.
\item $W^{[l]} \in \R$ \textsuperscript{\# of neurons in the next layer = j  x \# of neurons in the previous layer = k} is the weight matrix at layer $[l]$.
\item $b^{[l]} \in R$\textsuperscript{\# of units in next layer} is the bias vector in the $l^{th}$ layer.
\item $\hat{y} \in R^{n_{y}}$ is the predicted output vector. It can also be denoted as $a^{[L]}$ where $L$ is the number of layers in the whole network.
\item $g^{[l]}(x)$ is the $l^{th}$ activation function.  
\item $z^{[l]} = W_{x}x^{(i)} + b^{[l]}$ denotes the weighted sum of the input given to the $l^{th}$ layer before passing through the activation function.\\
\end{itemize}

@@

\subsection{Perceptrons}
\label{perceptron}
Perceptrons are the main components of neural networks. They were "developed in the 1950s and 1960s by the scientist Frank Rosenblatt, inspired by earlier work by Warren McCulloch and Walter Pitts" \cite{13}. Today, they are called "artificial neurons" or simply "neurons".\\
The output $a$ of a perceptron $j$ is a function $f$ of input $x=(x_{1}, ..., x_{n})$ weighted by a vector of weights $w_{}=(w_{1}, ..., w_{n})$, completed by a bias $b_{j}$ and associated to a non-linear activation function $g$:
\begin{equation}
\label{perceptron_equation}
a_{j} = f_{j}(x) = g((\sum_{k=1}^{n} x_{k} * w_{k}) + b_{j})
\end{equation}

Schematically speaking, a perceptron can be represented as on figure \ref{perceptron_model}. Each input is multiplied with its corresponding weight. The sum of this result then goes through a non-linear function, called "activation function". This activation function acts like a threshold that determines the proportion of the result that goes further in the network. There exist multiple activation functions (see \ref{activation_functions}). It is extremely important to use non-linear functions instead of a linear functions. In fact, the output of a perceptron is given as input to the others (see \ref{multilayer_perceptron}). Consequently, if only linear functions are used throughout the network, linear outputs are given as inputs to other linear functions. Since the composition of two linear functions is itself a linear function, assembling perceptrons to create neural networks of multiple layers would not make sense anymore.

@@

\subsection{Multilayer perceptrons}
\label{multilayer_perceptron}
A multilayer perceptron is a type of artifical neural networks. Du et al. \cite{23} define multilayer perceptrons as "feedforward networks with one or more layers of units between the
input and output layers" where "the output units represent a hyperplane in the space
of the input patterns."\\
A multilayer perceptron is composed of $L$ layers, each of them composed of $n^{l}_{h}$ perceptrons. The layers are organized in the following way:
\begin{itemize}
\item The input layer: It is the neural network entry point for the data. Generally speaking, the data is provided in the form of a matrix $X \in \R$ of size $(n_{x} \times batch\_size)$ with their corresponding labels $Y \in \R$ of size $(n_{y} \times batch\_size)$. The batch size defines the number of samples that will be given at the same time to the network and $n_{x}$ is the dimension of each sample. Moreover, $x^{(i)}$ is the $i^{th}$ sample represented as a column vector. The total number of training samples is given by $m$. Finally, $y^{(i)}$ is the output label for the $i^{th}$ example. For instance, suppose the number of samples is 100 and the batch size 32. In this situation, the network will be fed with 4 batches of sizes [32, 32, 32, 4] respectively.

\item The hidden layer(s): Hidden layers stand for all layers that are between the input layer and the output layer. Each of them has its own weights and biases (W, b), denoted by $W^{[l]} \in \R $ and $b^{[l]}$ respectively, where $W_{ij}^{l}$ corresponds to the weights associated with the connection between perceptron $j$ in layer $l$ and perceptron $i$ in layer $l+1$. By analogy, $b_{i}^{[l]}$ is the bias associated with perceptron $i$ in layer $l$. Weights and biases are the parameters to optimize in order to obtain the best mapping between the input and the output of the network (see Section \ref{training_a_neural_network} ). Before training the neural network, the weights can be randomly initialized or initialized with more sophisticated methods such as "Xavier initialization" or "Kaiming initialization"(see Section \ref{weight_initialization}).

\item The output layer: It is the last layer of the neural network. Its role is essential since it produces the prediction of the network for a given input. The prediction of a neural network is given by $\hat{y} \in R^{n_{y}}$ with $n_{y}$ representing the number of different labels. In a classification task, whose goal is to assign to each input a specific class, the $\hat{y}$ is usually the probability that the input belongs to each class. In that case, the \textit{softmax} activation function would be used at the end.
\end{itemize}
The advantage of organizing the weights, biases and inputs in matrices is due to the ability of modern CPUs/GPUs to quickly perform linear algebra computations. This way of structuring the network component is called "vectorization" which avoids using loops in the code, which would considerably slow down the computations.
Figure \ref{multilayer_perceptrons_figure} illustrates the concept of multilayer perceptrons. In this example, the total number of layers $L$ is equal to 3, the input size $n_{x}$ is equal to 4 and the number of units in in each layer is $n_{h}^{1}= 4$, $n_{h}^{2}=2$, $n_{h}^{3}=1$. The network contains the weights  $W^{(1)} \in \R^{2x3}$, $W^{(2)} \in \R^{1x2}$ and the biases  $b^{[1]}=2$, $b^{[2]}=1$.

@@

In other words, after each forward pass through the entire network, backpropagation performs a backward pass which aims at minimizing the cost function by adjusting parameters of the model. The way parameters are updated is defined by the gradients of the cost function with respect to each parameter of the network. The gradient of the cost function $J(x_{1}, x_{2}, ..., x_{m})$ at point $x$ is given by:
\begin{equation}
\frac{\partial J}{\partial x} = [\frac{\partial J}{\partial x_{1}}, \frac{\partial J}{\partial x_{2}}, ..., \frac{\partial J}{\partial x_{1}}]
\end{equation}
The gradient shows how much all parameters that constitute x need to change to minimize the function. In neural networks, the parameters of the cost function are all weight matrices $W^{[l]}$ and biases $b^{[l]}$. The computation of all these gradients relies on the "chain rule". In the case of weights, the chain rule is:
\begin{equation}
\frac{\partial J}{\partial w_{jk}^{l}} = \frac{\partial J}{\partial z_{j}^{l}} * \frac{\partial z_{j}^{l}}{\partial w_{jk}^{l}}
\end{equation}
Similarly, the chain rules has to be applied to the biases:
\begin{equation}
\frac{\partial J}{\partial b_{j}^{l}} = \frac{\partial J}{\partial z_{j}^{l}} * \frac{\partial z_{j}^{l}}{\partial b_{j}^{l}}
\end{equation}
Once the gradients of each parameter are computed, these parameters are updated. The weights update is described by the following equation:
\begin{equation}
W^{[l]} = W^{[l]} - \alpha * \frac{\partial J}{\partial W^{[l]}}
\end{equation}
The biases update corresponds to:
\begin{equation}
b^{[l]} = b^{[l]} - \alpha * \frac{\partial J}{\partial b^{[l]}}
\end{equation}
%The hyperparameter $\alpha$ of these two equations is called "learning rate". It determines the gradient's influence at each epoch and has to be manually tuned.
The "learning rate" $\alpha$ determines the influence that the gradient has at each epoch. It is an hyperparameter and has to be tuned manually.

@@

\subsection{Data}
In deep learning, data is essential. As seen previously, neural networks learn features from it. Therefore, data has to be handled carefully and in the right way. Usually, it is split into three different sets:
\begin{itemize}
\item The training set: This is the part of the dataset that is used to train the neural network (the weights and biases).
\item The validation set: This dataset is used to evaluate a trained model. Usually, the evaluation on the validation set is performed every $N$ epochs, where $N$ is a fixed number. The validation set needs to come from the same distribution as the training set but should contain exclusively unseen data. This last point is crucial since the validation set shows how well the neural network generalizes on unknown data. The validation set can also be used as indicator to decide when the training should be stopped in order to prevent "overfitting" which is the behaviour of a model that fits to the training set too closely and don't generalize well. In fact, if the validation loss continuously increases for a certain number of epochs, going on with the training will increase the overfit. This fact of interrupting the training earlier is called "early stopping".
\item The test set: This last part of the dataset is used to establish the final evaluation of the model. It also contains unseen data exclusively. 
\end{itemize}
Regarding the way these three sets are split, it mostly depends on the number of samples available. If the latter is big, the data is split into training and testing using the ratio 80/20. Then the remaining training samples are also split into training and validation using the ratio 80/20. On the contrary, if there are few data available, k-fold cross-validation is a good practice. This technique consists in splitting the entire dataset into k folds. One fold is picked as test set and the others are considered as training sets. The model is trained on training folds and tested on the test set. Then, another test set is picked and the same process is repeated until all possible tests set are picked. At the end of the process, the results of all test sets are averaged, which provides a good estimation of the model's performance. This technique is summarized on figure \ref{cross_validation}.

@@

\subsection{Hyperparameters tuning}
Hyperparameters denote parameters that cannot be directly learned from the data. That is the case for the learning rate, the batch size and the number of epochs that were described in previous sections. So, these parameters have to be manually tuned in order to find the best configuration (i.e the one that minimizes the cost function and that keeps an acceptable level of generalization).\\
Regarding the learning rate $\alpha$, its value has to be neither too large nor too small. A too large learning rate is recognizable by analyzing the training loss curve: if the loss is exploding or oscillating or if there is no improvement and it is stuck around a suboptimal local optima, the learning rate is too high and should be decreased. On the contrary, if the learning is very slow and the loss does not decrease or if the model is overfitting, it is a clear sign that it should be increased.
The learning rate can take a wide range of values. Consequently, the most used technique to find the optimal learning rate is simply the "trial and error", which consists in "trying widely different learning rates to determine the range of learning rates that need to be explored" \cite{28}. There also exists methods that, instead of keeping a fixed learning rate for the entire training, reduce it after each epoch (learning rate decay) or each time the loss on the validation set does not decrease (learning rate scheduling).\\
Batch size is another important hyperparameter to tune. Training a network with a small batch size "requires less memory, since the latter is trained using fewer samples" \cite{29}. Furthermore, "networks train faster because the weights update is done after each propagation" \cite{29}. Nevertheless, "the smaller the batch the less accurate the estimate of the gradient will be" \cite{29}. Indeed, due to the high weights update frequency, the gradient fluctuates much more than if was computed after a bigger number of samples.\\
Finally, the number of epochs during which the network is trained has to be carefully chosen. In fact, from a certain point in the training, neural networks don't learn anymore useful features in the data and start overfitting. This point corresponds to the moment where, the validation loss does not decrease anymore and starts to increase continuously. It is exactly at this moment that the training should stop. To achieve this goal, it is possible either to directly choose the right number of epochs or to use the so-called "early stopping" method, which stops the training as soon as the validation loss does not decrease for $N$ epochs.

@@

To perform a convolution, numerical input data and a "filter" are required. A filter can be seen as a $f$x$f$ numerical patch that moves across the entire input. It first moves horizontally until reaching the right-most border of the image. Then, it goes down a cell and starts from the border on the left-hand side. This process is repeated until the filter reaches the bottom-right corner of the image, which marks the end of the convolution. At each step, the dot product between the filter and the part of the input covered by the filter is computed. Figure \ref{fig:convolution} showcases a simple convolution which aims at finding vertical lines in a black and white image. For example, the output of the first step of the convolution (in red) is computed by evaluating the dot product between the blue filter and the red part of the input image: 
\begin{equation}
\begin{gathered}
\color{blue}(-1) \color{black}* \color{red}0 \color{black}+ \color{blue}2 \color{black}* \color{red}0 \color{black}+ \color{blue}(-1) \color{black}* \color{red}0
\\ \color{black}+ \\ 
\color{blue}(-1) \color{black}* \color{red}0 \color{black}+ \color{blue}2 \color{black}* \color{red}0 \color{black}+ \color{blue}(-1) \color{black}* \color{red}1 
\\ \color{black}+ \\ 
\color{blue}(-1) \color{black}* \color{red}0 \color{black}+ \color{blue}2 \color{black}* \color{red}0 \color{black}+ \color{blue}(-1) \color{black}* \color{red}1 
\\ \color{black}= -2
\end{gathered}
\end{equation}

@@

Different parameters can change the way a convolution behaves. First of all, the previous example relied on a filter moving by respectively one cell to the right and to the bottom. In this case, the so-called "stride" is equal to 1. Other applications could rely on a bigger stride. Furthermore, the previous example reduced the output size of $3$x$3$ in proportion to the initial input size of $5$x$5$. To influence the output size, a padding can be added to the outside of the input image, usually filled with 0s. Three ways of padding images are commonly used as shown on figure \ref{fig:convolution_padding}:
\begin{itemize}
	\item \textbf{Valid:} The input image is not padded. This means that the filter only goes through existing pixel values, which makes the output size smaller than the input size.

	\item \textbf{Same:} The input image is padded in a way that makes the output size the same as the input size.

	\item \textbf{Full:} The input image is padded so that, in the first step of the convolution, only the bottom-right cell of the filter overlays the pixel value of the image, the rest overlaying padding cells. This makes the output size larger than the input size. 
\end{itemize}
